The CMS Data Aggregation System

### 2010-05-04 doc/paper.tex
Made 1/2 pass through the paper, up to 4.1 chapter. A lot of changes. Simon please have a look.
git-svn-id: svn+ssh://svn.cern.ch/reps/CMSDMWM/DAS/trunk@632 4525493e-7705-40b1-a816-d608a930855b

Meta-data plays signiﬁcant role in a large modern enterprises, 
research experiments and digital libraries where it comes from different 
sources and distributed in a variety of forms and digital formats. 
It is organized and managed by constantly evolving software using 
both relational and non-relational data sources. 
Even though we can access it via set of data-services, twikies, blogs, etc.,
there is no coherent way for information discovery from different sources
in heterogeneous, distributed environment with different security policies enforced.

Here we discuss a new data aggregation system which consumes, 
indexes and delivers information from different relational and 
non-relational data sources to answer cross data-service queries 
and explore meta-data associated with petabytes of experimental data. 
We combine simplpicity of keyword-based search with precision of RDMS
under a new system where aggregated information being collected from various sources 
allowing end-users to place dynamic queries, get precise answers and 
trigger information retrieval on demand. Such close to real-time system 
based on studies  of use cases of the CMS particle physics experiment at 
the LHC which uses a large scale, distributed computing system to motivate the work.

@@

\section{Introduction}
The CERN, the European Organization for Nuclear Research, plays a leading
role in fundamental studies of physics. It is also known as a place where
World Wide Web was born to help researches share information among each other
via concept of hypertext.
Today, the Large Hadron Collider (LHC) at CERN is marking a new era of High Energy
Physics (HEP), promising to deliver a few PB of data each year. 
At this scale scientists are facing with a new set of problems in an area of
``information discovery'' within heterogeneous, distributed environment.
The data and associated meta-data are produced in variety of forms and digital formats.
They are stored and retrieved from relational and non-releational data-sources, such as 
RDMS systems, document oriented databases, blogs, twikies, file systems and
customized applications, etc. Its access is restricted by broad variety of security 
policies enforced by individual sites and community in general.
Working in such environment is not easy and require a lot of knowledge, but users
(in this case physicists) are always looking for simple, intuitive and flexible
tool to look-up their desired data. A well-known solutions, such as web-services,
blogs, twikies and search engines targeted and tighted to a specific data
sources and end-users are left with a manual task to bookkeep and relatate information
from them.

Here we present a work on Data Aggregation System (DAS) designed for
CMS High-Energy Experiment operated at LHC which provide
ability to query, search and aggregate information from different 
data-services as well as serve them as a caching layer, 
preserving their security policies.

We organize
our discussion as following. In section \ref{RelatedWork} we discuss
related work in a domain of keyword search over relational data sources.
The section \ref{DataModel} describes CMS experiment and its data model. In section
\ref{DAS} we outline architecture of the DAS system, including discussion of its
various components. Finally our results are summarized in section \ref{Results}.

@@

\section{Related Work\label{RelatedWork}}
Even though the idea of querying relation databases via keyword based search
algorithms is not knew it is still under significant activity in computer
science domain. A few alternative solutions has been proposed to address this issue
in last several years. The federated DB \cite{FedDB} by IBM unifies data coming 
from different RDMS systems into federated DB where SQL queries can be placed to 
search desired data. The other approaches \cite{DBXplorer, QueryAnswer}, influenced by 
great success of search engines, explored ability to use keyword search algorithms 
over relataional RDMS by indexing DB content. In former case, a problem with understanding 
of underlying schema and imposing relational conditions in a query still exists, while 
in later queries can be expressed via simple keyword based search, even though the exact match
of provided keywords is expected. Although those solutions exist,
with respect to HEP and other domains they are not sufficient.
In a case of federeated database exact knowledge of schema is required. This means
that end-users should be aware of database internal sctructure in order to use SQL.
While the proposed keyword search based systems
work in confined environment and do not provide ability to use condition expressions
over relational data. For instance, as quoted in \cite{QueryAnswer}, if someone asks about 
{\it Woody Allen} a set of database records can be found to satisfy this request.
Such approach cannot be used to answer the logical questions, such as 
{\it find me total number of files whose size more then 20 but less then 100 taken from january
to february of 2009} which is easy to address using SQL syntax.

To address those issues
an attempt to build a simple, intuitive and flexible query language was introduced
in \cite{DBS-QL, AMI}.
%\footnote{The ATLAS experiment at CERN, has been developed
%AMI web portal \cite{AMI} with similar functionality.}
It represented a power of SQL while
hiding underlying relational schema from the end-users. As a results
a human questions were intuitively mapped into simple queries. For example,
the question
{\it I'm looking for files who contain data taken on certain date and located at
particular site} was represented as simple as \cite{DBS-QL}
\begin{verbatim}
find file where date 2009-01-02 11:59 CET and site = T2
\end{verbatim}
Such approach not only address common use cases within HEP community, but also
represented an intuitive mapping between mental model of our users and 
underlying database back-end.

We wanted to expand this approach further and apply it 
to handle cross data-service queries from broad variety 
of relational and non-relational data-sources within 
our distributed heterogeneous environment.

\section{CMS data model\label{DataModel}}
The CMS, Compact Muon Solenoid, experiment \cite{CMS} 
is one of the two large general-purpose particle physics detectors built on 
the proton-proton Large Hadron Collider (LHC) at CERN in Switzerland and France. 
It is designed to explore the frontier of High Energy Physics and provide physicsists
ability to look at conditions which were present at early stage of our Universe.
More then 3000 physicists from 183 institutions reprepresenting almost 
38 countries are involved in design, construction and maintenance of the experiment.

The CMS distributed computing and data model \cite{CMSDataModel} 
is designed to process
and efficiently manage a few PBs of data expected to be produced each year
at LHC. The computing resources provided by members of CMS
collaboration are geographically distributed, 
interconnected via high throughput networks and operated by means 
of Grid software. The model allows to cover broad variety of
hardware, mass and storage elements and configuration of the
clusters. To accomodate efficient data processing CMS uses
a multi-Tier distributed model, where specific tasks of data taking,
processing, archival and distribution are assigned to each tier based
on CMS data model. For example, the Tier-0 center at CERN is responsible
for archiving the data coming out from the detector, prompt first pass reconstruction
and data distribution to Tier-1 centers. The 7 Tier-1 centers
located in France, Germany, Ialy, Spain, Taiwan, United Kingdom and United States
keep a portion (copy) of the data delivered by Tier-0 for further processing.
They provide storage and CPU power for high priority analysis jobs.
The Tier-2 centers, located at more then 50 sites around the world,
are dedicated for user analysis tasks and production of simulated data.

A broad variety of data-services being designed and developed to
maintain detector and production operations, including detector
conditions databases, data location and bookkeeping services,
data transfer and job monitoring tasks. Even though majority of them
are located at CERN, it was never been a requirement in CMS computing
and data model. For instance, the production teams operated at Tier-1,2
centers set up local services for data bookkeeping and operational
tasks. Based on their resources the choice of back-end DB was
up to the site managers. Therefore the CMS software were designed to support different
DB back-ends and provide tools for migrating data across them.

Once such conglomerate of data-services starts operating an obvious
question arise: how to find out desired information across multiple data-services
in our distributed environment? Even though individual data-services were designed
to answer specific questions about data they serve, the ability to search and relate
information among them was tedious human task. The growing amount of information
and desire to make cross-service queries force us to design and develop a new
type of system, the Data Aggregation System.

\section{Data Aggregation System\label{DAS}}
The design of DAS system was based on previous studies of CMS Data 
Bookkeeping System (DBS) use cases \cite{DBS, DBS07}. We carefully analyzed user
requests and queries placed to DBS, their patterns, frequencies and latency. As a result
we created a data discovery service \cite{DD} which came through several iterations
of design choices. At the end the end-user interface was based on DBS 
Query Language (DBS-QL)\cite{DBS-QL} which supports dynamic queries against
DBS back-end. The DBS-QL \cite{DBS-QL} syntax was based on SQL query language with one 
important difference, we took away any knowledge about relations of underlying schema
from end-users. It was achieved by usage of Dijkstras shortest path algorithm 
to establish necessary table joins based on provided information in user queries and
knowledge of foreign key relationships from auto-loaded database schema.
The DBS-QL keys exposed to end-users were choosen based on jargon used by
physicists. They're internally mapped into database entitites and attributes, 
e.g. {\it run} DBS-QL key refers to run number attribute of run summary table.
A quick adoption and wide usage of DBS-QL gave us confidence of choosen approach 
and provide a basis for building DAS system.

As was pointed out in \cite{Arms} a mixed content and 
mixed meta-data and meta-data consistency should be considered as a whole in design 
of the system to successful information discovery. 
Starting from this ground we designed DAS system as an
additional layer on top of the existing data-services
within CMS computing infrastructure by imposing the following set of requirements:
\begin{itemize}
\item support flexible and intuitive Query Language (QL) to support
both keyword and condition queries;
\item pluggable interface to existing and up-coming data-service APIs;
\item support heterogeneous software environment and distributed nature of data-services;
\item preserve security policies of individual data-services;
\item be flexible to adjust to dynamic nature of data-service APIs, 
due to continuously changing detector and software environments;
\item retrieve data on demand and aggregate them if necessary across
multiple data-services;
\item be transparent to data-service's implementations and data formats;
\item support legacy data-services, applications and APIs.
\end{itemize}

%\begin{figure}[htb]
%\centering
%\includegraphics[width=150mm]{DAS_architecture.pdf}
%\caption{
%DAS architecture.
%}
%\label{DAS_arch}
%\end{figure}
\begin{figure}[htb]
\centering
\includegraphics[width=150mm]{DAS_Cache_and_Analytics.pdf}
\caption{
DAS architecture diagram. It consists of DAS cache server, mapping and analytics DBs.
Once query was placed to DAS the associated
data-service APIs were invoked, triggering DAS cache population. Once data become
available and aggregated in DAS cache users were able to see the results. This on demand
system, where data were retrieved and store into cache from
back-end data services, was supplemented by DAS robots (unix daemons) who
were consulting DAS analytics database to enable data population into DAS
cache for most common queries.
}
\label{DAS_cache}
\end{figure}

\noindent
The DAS architecture is shown in Fig. \ref{DAS_cache}. It consists of the
following components:
\begin{itemize}
\item core library with support of pluggable modules for data retrieval;
\item caching layer to store data-service output and aggregated results;
\item request service to handle user requests;
\item mapping DB to keep information about data-service APIs, their
notations and mapping to DAS;
\item analytics DB with query statistics to enable pre-fetching 
strategies.
\end{itemize}

The DAS core communicates with different data-services and retrieves
data on demand. This was done by invoking appropriate data-service APIs
upon user requests. All user queries were written to Analytics DB
for further analysis. The data-service output were re-mapped into
DAS notations and, if necessary, aggregated inside of DAS cache.
Since we didn't know a-priory what kind of data users will ask, 
what kind of queries they will place against the system and what
level of aggregation will be applied the choice
of RDMS as DAS back-end was problematic\footnote{In addition DAS 
didn't impose any requirements on transactions and persistency of the data.} 
and we look around for other options. Through the analysis of available IT solutions, 
such as file-based and memory caches, 
key-value databases, documented-oriented databases we made our choice in favor 
of the last technology. Among of them we evaluated CouchDB \cite{CouchDB} and 
MongoDB \cite{MongoDB}. Both systems provide schema free document
storage, replication and fail-over features, but our choice in favor of 
MongoDB was obvious because of its support of dynamic queries, 
full indexies, including on inner objects and embedded arrays,
and auto-sharding. Our prelimiary benchmarks shown that it can sustain
the desired load and size for our meta-data information. We use MongoDB as a back-end 
for the three main components: Mapping, Analytics and DAS cache databases.

DAS Mapping DB holds mapping between data-service APIs and DAS keys. 
Upon identification of APIs which will participate in DAS activity 
we store to Mapping DB the API name, input parameters, their type and accept 
values as regular expression patterns as well as mapping between data-service 
notations and DAS keys. It allows us to map user input query parameters into
data-service API metrics as well as map back data-service output into DAS notations.
For instance, the DBS system \cite{DBS}
used {\it logical\_file\_name} notation, while in PhEDEx
\cite{PhEDEx}
the same entity was named as {\it lfn}. Since both of them were referred to
the logical file name, we map both notations into DAS key {\it file}.\footnote{We
also deal with physical file names, but still used {\it file} to refer to them in
DAS notationis, since simple regular expressions were used to identify the
meaning of the {\it file} keyword from the provided value.}
For each service we provide a mapping from its own data-format (JSON, XML, etc.) into
common data-format used in DAS.\footnote{We used JSON as a DAS data format.}
We also store a mapping between DAS keys and data-service API names to
establish appropriate calls to data-services upon provided DAS query. For example,
when user specifies {\it file} in their DAS query, we invoke DBS 
{\it listFiles} and PhEDEx {\it fileReplicas} APIs, respectively.

The DAS Analytics DB plays a special role. It collects information
about user requests placed to the system. Each request was recorded. Upon its
decomposition into set of selection keys and conditions we also record
API name and API input parameters into Analytics database and 
keep updating their counter information for repeated calls. 
This allows to keep track of frequency of API calls and establish 
pre-fetch strategies for most common queries.

The DAS caching system was used to store data-service results and aggregated
information in form of DAS records. As it shown in figure \ref{DAS_cache} 
the DAS cache server updates Analytics DB, requests and retrieve data from DAS cache
and aggregate data over there. The independent set of daemons, DAS robots, are
constantly monitor Analytics DB for most popular requests and pre-fetch most
common data into DAS cache.

Each user request placed to the system was parsed and set of selection
and condition keys were identified. They were mapped into set of
data service APIs and their input parameters. DAS
cache was checked for a similar set of selection and condition
keys which can be used earlier.\footnote{Since DAS doesn't hold
any data by default we cannot rely on data being present in a cache.
Moreover if we look-up data first, we cannot tell if it was complete
set or not, since earlier queries may retrieve only portion of the requested data.}
If the superset was found a data were
retrieved from the cache, otherwise a data-service APIs were
invoked and the results were collected and aggregated in DAS cache.

%Now we present overall logic of DAS workflow and details of individual components.
%Upon request to DAS the following steps were executed: 
%\begin{itemize}
%\item parse input DAS query
%\item map provided selection and condition keys into data-service APIs and
%input parameters
%\item look-up similar queries in DAS cache 
%\item invoke data-service APIs
%\item parse output results by mapping output notation into DAS keys
%\item store and aggregate them into DAS raw-cache
%\end{itemize}

Apart from previous DBS-QL syntax we decided to fully support free text-based queries for end-users.
All of them were transformed into MongoDB query syntax, who was used internally in
DAS core layer. It represents a set of selection keys and conditions as a JSON document.
For example, a query
{\it dataset site=T1\_CH\_CERN}
where transformed into 
{\it \{"fields" : ["dataset"], "spec" : \{"site" : "T1\_CH\_CERN"\}\}}.
Such syntax was perfect fit for us, since it allows to store all queries
as JSON documents into Analytics DB for further analysis. It is worthwhile to note that
MongoDB query syntax is very reach. It supports advanced queries with 
boolean expressions, conditional operators, regular expressions, etc. providing
a real power for our end-users.

@@

Based on this results we measured that we can achieve 7000 docs/sec rate
for insertion, 3400 docs/sec for aggregation and 8500 docs/sec for reading operation
in DAS, which is suitable for our 
needs.\footnote{We've been satisfied with such performance,
even though our measurements shown that it still can be speed up by few times
using C++ DB driver instead of python one used in this prototype system.}

@@

\section*{References}
\begin{thebibliography}{9}
\bibitem{FedDB}
L. Haas, E. Lin,
``IBM Federated Database Technology'', \\
http://www.ibm.com/developerworks/data/library/techarticle/0203haas/0203haas.html

\bibitem{DBXplorer}
Sanjay Agrawal, Surajit Chaudhuri, Gautam Das: DBXplorer: A System for
Keyword-Based Search over Relational Databases. ICDE 2002: 5-16

@@

>>>>

Meta-data plays signiﬁcant role in a large modern enterprises, 
research experiments and digital libraries where it comes from different 
sources and distributed in a variety of forms and digital formats. 
It is organized and managed by constantly evolving software using 
both relational and non-relational data sources. Even though we can apply
information retrieval approach to non-relation data sources,
we can't do so for relational ones, where information can be accessed via
pre-established set of data-services.
%Even though we can access it via set of data-services, twikies, blogs, etc.,
%the information discovery across them is still a manual task.
% which prevent
%information discovery
%there is no coherent way for information discovery from different sources
%in heterogeneous, distributed environment with different security policies enforced.

Here we discuss a new data aggregation system which consumes, 
indexes and delivers information from different relational and 
non-relational data sources to answer cross data-service queries 
and explore meta-data associated with petabytes of experimental data. 
We combine simplpicity of keyword-based search with precision of RDMS
under the new system. The aggregated information being collected from various sources,
allowing end-users to place dynamic queries, get precise answers and 
trigger information retrieval on demand. Such close to real-time system 
based on studies  of use cases of the CMS particle physics experiment at 
the LHC where we used a large scale, distributed computing system 
to motivate the work.

@@

\section{Introduction}
The European Organization for Nuclear Research, known as CERN, plays a leading
role in fundamental studies of physics. But it is also known as a place where
many innovations in the area of computer science were developed, e.g. World Wide Web.
Today, the Large Hadron Collider (LHC) at CERN is marking a new era of High Energy
Physics (HEP), promising to deliver a few PB of data each year. 
At this scale the ``information discovery'' within heterogeneous, distributed 
environment become a key ingredient of successful data analysis in HEP.
%scientists are facing with a new set of problems in an area of
%``information discovery'' within heterogeneous, distributed environment.
The data and associated meta-data are produced in variety of forms and digital formats.
They are stored and retrieved from relational and non-releational data-sources, such as 
RDMS systems, document oriented databases, blogs, twikies, file systems and
customized applications, etc. 
%Its access is restricted by broad variety of security 
%policies enforced by individual sites and community in general.
Working in such environment requires a lot of expertise and our users
(in this case physicists) are always looking for simple, intuitive and flexible
tool(s) to look-up their desired data. A well-known solutions, such as data-services
and search engines tighted to a specific data sources and end-users are left 
with a manual task to bookkeep and relatate information from them.

Here we present a work on Data Aggregation System (DAS) designed for
CMS High-Energy Experiment at LHC which provides
ability to query, search and aggregate information from different 
data-services and trigger information retrieval on demand.

The rest of this paper is organized as following. 
In section \ref{RelatedWork} we discuss related work in a domain of 
keyword search over relational data sources.
Section \ref{DataModel} describes CMS experiment and its data model. In section
\ref{DAS} we present architecture of the DAS system, including discussion of its
various components. Finally, our results are summarized in section \ref{Results}.

@@

\section{Related Work\label{RelatedWork}}
Even though the idea of keyword queries over the relational database(s)
%querying relation databases via keyword based search
%algorithms 
is not knew it is still under discussion in computer science domain.
A few alternative solutions has been proposed to address this issue
in last several years. In \cite{DBXplorer} the conjunctive keyword queries,
i.e. retrieval of only documents that contain all query keywords, has
been discussed. It requires that all keywords are matched in a row from
single table or joined tables. While \cite{QueryAnswer} make a step forward
by considering information around matched values and presents customized answers to
end-users over the generated new database. Those and other works\footnote{
For more information about
keyword based search over relational databases we redirect users to reference in
\cite{DBXplorer, QueryAnswer}.} 
are based on some kind of the graph over the underlying database and
moreover use a generated database out of the pristine one.
The proximity of the results in those techniques 
are mostly confined to the text-based values stored in a database, since the
keyword based search with numerical values 
need the precise knowledge of the related entity in a schema. For example, typing
100 can lead to inappropriate matches, such as row ids, and in order to be
answered correctly it requires additional keyword, clarifying its context, 
which should be matched with particular
entity (table.column) in the underlying database.
Therefore such approaches cannot be used to answer the logical questions, such as  
{\it find me total number of files whose size more then 20 but less then 
100 taken from january to february of 2009} which can be easily accomplished
using SQL syntax.

To address those limitations an attempt to build a simple, intuitive and 
flexible query language was introduced in \cite{DBS-QL}\footnote{A similar but less
sophisticated approach was done in \cite{AMI}}.
It represents a power of SQL while hiding underlying relational schema from 
the end-users. As a results a human questions were intuitively mapped into 
simple queries. For example, the question
{\it I'm looking for files who contain data taken on certain date and located at
particular site} was represented as simple as \cite{DBS-QL}
\begin{verbatim}
find file where date 2009-01-02 11:59 CET and site = T2
\end{verbatim}
This solution represents a common use cases within HEP community and provide
intuitive mapping between mental model of end-users and underlying database back-end.

At the same time the data aggregation is commonly addressed in the context of
federated databases, see for example \cite{FedDB}. It unifies data coming 
from different RDMS systems into federated DB where SQL queries can be placed to 
search desired data. Even though it would be interesting to apply technique 
discussed in \cite{DBS-QL} to a federated database it still requires 
an access to the underlying database schema. This obviously creates a problem
working in distributed heterogeneous environment where different security policies
are in place.

To avoid those limitations we decided to build a keyword-search based system around
existing data-services whose nature and access policies are defined by developers.
Keeping in mind that meta-data can come in different forms and data formats
we used a document-oriented database back-end with full support of query language.
The key ingredients were the setup of the mapping and analytics services which provide
necessary glue across existing API/notations and user queries. This approach 
allows us to perform the data aggregation across various data-services, achieve
precise answers using keyword-based queries
and organize information retrieval on demand.
%to store meta-data information 
%We wanted to expand this approach further and apply it 
%to handle cross data-service queries from broad variety 
%of relational and non-relational data-sources within 
%our distributed heterogeneous environment.

\section{CMS data model\label{DataModel}}
The CMS, Compact Muon Solenoid, \cite{CMS} 
is one of the two large general-purpose particle physics detectors built on 
the proton-proton Large Hadron Collider (LHC) at CERN in Switzerland and France. 
It is designed to explore the frontier of High Energy Physics and provide physicists
ability to look at conditions presented at early stage of our Universe.
More then 3000 physicists from 183 institutions reprepresenting almost 
38 countries are involved in the design, construction and maintenance of the experiment.

The CMS distributed computing and data model \cite{CMSDataModel} 
is designed to process
and efficiently manage a few PBs of data expected each year
in CMS operation at LHC. The computing resources provided by members of CMS
collaboration are geographically distributed, 
interconnected via high throughput networks and operated by means 
of Grid software. The model allows to cover broad variety of
hardware, mass and storage elements and configuration of the
clusters. To accomodate efficient data processing CMS uses
a multi-Tier distributed model, where specific tasks of data taking,
processing, archival and distribution are assigned to each tier based
on CMS data model. For example, the Tier-0 center at CERN is responsible
for archiving the data coming out from the detector, prompt first pass reconstruction
and data distribution to Tier-1 centers. The 7 Tier-1 centers
located in France, Germany, Ialy, Spain, Taiwan, United Kingdom and United States
keep a portion (copy) of the data delivered by Tier-0 for further processing.
They provide storage and CPU power for high priority analysis jobs.
The Tier-2 centers, located at more then 50 sites around the world,
are dedicated for user analysis tasks and production of simulated data.

A broad variety of data-services being designed and developed to
maintain detector and production operations, including detector
conditions databases, data location and bookkeeping services,
data transfer and job monitoring tasks. Even though majority of them
are located at CERN, it was never been a requirement in CMS computing
and data model. For instance, the production teams operated at Tier-1,2
centers set up local services for data bookkeeping and operational
tasks. Based on their resources the choice of database back-end was
up to the site managers. Therefore the CMS software were designed to support different
DB back-ends and provide tools for migrating data across them.

Once such conglomerate of data-services starts operating an obvious
question arise: how to find out desired information across multiple data-services
in our distributed environment? Even though individual data-services were designed
to answer specific questions about the data they serve, the ability to search and relate
information among them was always a tedious human task. The growing amount of information
and desire to make cross-service queries force us to design and develop new
Data Aggregation System.

\section{Data Aggregation System\label{DAS}}
The design of the DAS system was based on previous studies of CMS Data 
Bookkeeping System (DBS) \cite{DBS, DBS07}. We carefully analyzed user
requests and queries placed into the DBS, their patterns, frequencies and latency. 
As a result we created a data discovery service \cite{DD}. Originally it used MVC
architecture and later, being re-factored as pure presentation layer on top of 
the DBS Query Language (DBS-QL) \cite{DBS-QL} with support of dynamic queries against
DBS back-ends. The DBS-QL \cite{DBS-QL} syntax was based on SQL query language, where 
we took away relations and details on underlying schema from end-users. 
This was achieved by usaging Dijkstras shortest path algorithm 
to establish necessary table joins based on provided information in user queries and
knowledge of foreign key relationships from auto-loaded database schema.
The DBS-QL keys exposed to end-users were choosen based on a jargon used by
physicists. They're internally mapped into database entitites and attributes, 
e.g. the {\it run} DBS-QL key refers to run number attribute of run summary table.
A quick adoption and wide usage of DBS-QL gave us confidence in choosen approach 
and provide a basis for building DAS system.

As was pointed out in \cite{Arms} a mixed content and 
mixed meta-data and meta-data consistency should be considered as a whole in design 
of the system to successful information discovery. 
Starting from this ground we designed DAS system as an
additional layer on top of the existing data-services
within CMS computing infrastructure by imposing the following set of requirements:
\begin{itemize}
\item support of keyword based search queries with ability to use condition operators;
\item pluggable interface to existing and up-coming data-service APIs;
\item support heterogeneous software environment and distributed nature of the data-services;
\item preserve security policies of individual data-services;
%\item be flexible to adjust to dynamic nature of data-service APIs, 
%due to continuously changing detector and software environments;
\item retrieve data on demand and aggregate them if necessary across
multiple data-services;
\item be transparent to data-services implementations and their data formats;
%\item support legacy data-services, applications and APIs.
\end{itemize}
\begin{figure}[htb]
\centering
\includegraphics[width=150mm]{DAS_Cache_and_Analytics.pdf}
\caption{
The DAS architecture diagram. 
Input query where placed into analytics DB and mapped into set of
data-service APIs and their parameters. If superset of the query has been
found in DAS cache, the appropriate results were delivered back to users.
Otherwise a data-service APIs were invoked, triggering DAS cache population. 
Once data become available and aggregated in DAS cache users were able 
to see the results. This on demand system was supplemented by DAS robots 
(unix daemons) who were consulting DAS analytics database to enable data 
population into DAS cache for most common query requests.
}
\label{DAS_cache}
\end{figure}

\noindent
Due to this set of requirements, we rule out the choice of RDMS as DAS back-ends
for several reasons. We agreed to support dynamic queries, data
aggregation on demand and pluggable data-services. Therefore a choice 
of concrete schema for data storage was problematic. Moreover we didn't impose
any transactions and data persistency in DAS. This allows us to look 
for alternative IT solutions.
Through the analysis of available options, such as a file-based and memory caches, 
key-value databases, documented-oriented databases we made our choice in favor 
of the last technology. Among them we evaluated CouchDB \cite{CouchDB} and 
MongoDB \cite{MongoDB}. Both systems provide schema free document
storage, replication and fail-over features, but our choice in favor of 
MongoDB was obvious, due to its support of dynamic queries, 
full indexies, including inner objects and embedded arrays,
and auto-sharding. Our prelimiary benchmarks shown that it can sustain
the desired load and size for our meta-data information. We used 
document-oriented database, MongoDB, as a back-end for the three 
main DAS components: Mapping, Analytics and DAS cache databases, 
which will be discussed next.

The DAS architecture is shown in Fig. \ref{DAS_cache}. It consists of the
following components:
core library with support of pluggable modules for data retrieval;
caching layer to store data-service output and aggregated results;
request service to handle user requests;
mapping DB to keep information about data-service APIs, their
notations and mapping to DAS;
analytics DB for query analysis and identification of pre-fetch 
strategies.

The DAS core communicates with different data-services and retrieves
data on demand. This was done by invoking appropriate data-service plugin
upon the user request. Each plugin was responsible for invoking
data-service APIs and parsing retrieved data into the DAS data format
suitable for storage, JSON. Although the majority of data-services looks alike
and their logic was abstracted, including parsers, there were a few
cases where special care was required. For instance,
we dealt with data-services which were only accessible via private network.
Therefore, a way to propagate user credentials was required, leading
to specific plugin implementation.

%All user queries were written to Analytics DB for further analysis, while
%mapping between data services were kept in Mapping DB. The data-service 
%output were re-mapped into DAS notations and, if necessary, aggregated 
%inside of the DAS cache.

DAS Mapping DB holded the mapping between data-service APIs and DAS keys exposed to
the users. It created a necessary glue between data-services and identifies
how their data can be related to each other. Basically Mapping DB was positioned 
as a {\it registration} service with ability to translate notations among data-services.
Upon identification of APIs which will participate in DAS activity 
we stored their names, input parameters, together with their type and accept 
values as regular expression patterns into the Mapping DB. 
Also we stored the mapping between data-service notations and DAS keys. 
For instance, the CMS DBS system \cite{DBS}
used {\it logical\_file\_name} notation to identify the name of the
logical file, while the same entity was called {\it lfn} in CMS PhEDEx system \cite{PhEDEx}.
Since both of them were referred to the logical file name, 
we map both notations into DAS key {\it file}.\footnote{We
also deal with physical file names, but still used {\it file} to refer to them in
DAS notations, since simple regular expressions were used to identify the
meaning of the {\it file} keyword from the provided value.}
Such keys, e.g. {\it file}, were exposed to the end-users as pre-defined keywords to
be used in search queries. For instance, end-users were able to type {\it file} in
order to get all file names known to DAS via data-services as well as
provide a concrete patterns associated with a file, e.g. {\it file=/abc*}.
Under the hood, the appropriate data-service APIs were looked up along with appropriate
parameters to answer those questions. For example,
in aforementioned example, using {\it file} in DAS query cause invokation of DBS 
{\it listFiles} and PhEDEx {\it fileReplicas} APIs, respectively.
This approach gave power to end-users to express the meaning of their queries, especially
when they deal with numbers and achieve accuracy of results from DAS. 
For instance, users were able to type {\it run > 10 and run < 100}
to specify a range of run numbers to be looked up in DAS cache. At the same time
a full free text based keyword search is under development within MongoDB.

The DAS Analytics DB played a special role. It collected information
about user requests placed to the system. Each request was recorded. Upon its
decomposition into set of selection keys and conditions we also recorded
API name and input parameters into the Analytics database and 
keep updating their counter information for repeated calls. 
This allows to keep track of frequency of API calls and establish 
pre-fetch strategies for most common queries.

The DAS caching system was used to store data-service results and aggregated
information in form of DAS records. As it shown in figure \ref{DAS_cache} 
the DAS cache server updates Analytics DB, requests and retrieve data from DAS cache
and aggregate data over there. The independent set of daemons, DAS robots, were
constantly monitor Analytics DB for most popular requests and pre-fetch most
common data into DAS cache.

Each user request placed to the system was parsed and set of selection
and condition keys were identified. They were mapped into set of
data service APIs and their input parameters. DAS
cache was checked for a queries with a similar set of selection and condition
keys used earlier.\footnote{Since DAS didn't hold
any data by default we cannot rely on data being present in a cache.
Moreover if we look-up data first, we cannot tell if it was complete
set or not, since earlier queries may retrieve only portion of the requested data.}
If the superset was found a data were
retrieved from the cache, otherwise a data-service APIs were
invoked and the results were collected and aggregated inside of DAS cache.

Apart from previous DBS-QL syntax we decided to drop its syntax and allow users
express their queries in a free text-based form. Such queries were transformed 
into MongoDB query syntax, who was used internally in
DAS core layer. It represents a set of selection keys and conditions as a JSON document.
For example, a query
{\it dataset site=T1\_CH\_CERN}
where transformed into 
{\it \{"fields" : ["dataset"], "spec" : \{"site" : "T1\_CH\_CERN"\}\}}.
Such syntax was perfect fit for us, since it allows to store all queries
as JSON documents into Analytics DB for further analysis. It is worthwhile to note that
MongoDB query syntax is very reach. It supports advanced queries with 
boolean expressions, conditional operators, regular expressions, etc. providing
a real power for our end-users.

@@

Based on this results we measured that we can achieve 7000 docs/sec rate
for insertion, 3400 docs/sec for aggregation and 8500 docs/sec for reading operation
in DAS, which is suitable for our 
needs.\footnote{We were satisfied with performance of the DAS prototype,
even though it throughput can be increased by few times
using C++ DB driver instead of python one used in this system.}

@@

\section*{References}
\begin{thebibliography}{9}
\bibitem{DBXplorer}
Sanjay Agrawal, Surajit Chaudhuri, Gautam Das: DBXplorer: A System for
Keyword-Based Search over Relational Databases. ICDE 2002: 5-16

@@

\bibitem{FedDB}
L. Haas, E. Lin,
``IBM Federated Database Technology'', \\
http://www.ibm.com/developerworks/data/library/techarticle/0203haas/0203haas.html
